\texttt{ensmallen}: a generic C++ library for fast optimization

### 2018-10-14 doc/mlsys_paper/paper.tex
improved titles
sharpening and defluffing the intro

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  Almost all machine learning problems can be boiled down to the
simple optimization

\begin{equation}
\operatorname{argmin} f(x)
\end{equation}

\noindent for some objective function $f(\cdot)$; in some cases, these objective
functions may have some special structure or some constraints.

Often, these optimizations can be computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data. %TODO: cite
Even popular machine learning models such as logistic regression % TODO: cite
have training times mostly dominated by an optimization.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms, which is
what drives us to introduce {\tt ensmallen}, a generic C++ optimization toolkit
that contains a wide variety of optimization techniques for different types of
objective functions.  Through the use of template metaprogramming, {\tt
ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take advantage of the types of compiler optimizations that {\tt
ensmallen} can, and few offer the kind of robust support for different types of
objective functions that {\tt ensmallen} does.  For instance, deep learning
libraries like Caffe, % TODO: cite
PyTorch, % TODO: cite
and TensorFlow % TODO: cite
each contain a variety of optimization techniques; but these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Some other machine learning
libraries, such as {\tt scikit-learn} % TODO: cite
% TODO: can we find another library to put here too?  Shark maybe?
contain optimization algorithms but not in any coherent or reusable framework.
On the other hand, many programming languages have higher-level packages for
mathematical optimization; {\tt scipy.optimize} % TODO: check and cite
is widely used in the Python community, and MATLAB's {\tt TODO()}
% TODO: what's the name of that MATLAB function? or one of them?
% TODO: cite
has been available and used for many decades.  However, these implementations
don't tend to be suitable for modern machine learning tasks---for instance,
computing the full gradient of the objective function may not be feasible
because there are too many data points.

In this paper, we describe the functionality of {\tt ensmallen} and the types of
problems that it solves.  We discuss the mechanisms by which {\tt ensmallen} is
able to provide both computational efficiency and ease-of-use, and then show a
few examples that use the library.  Then, we show performance comparisons with
other optimization libraries.

{\tt ensmallen} is available for download from the website at
\url{https://www.ensmallen.org}, and is currently maintained on Github at
\url{https://github.com/mlpack/ensmallen}.  Originally, {\tt ensmallen} was
developed as the optimization infrastructure of {\tt mlpack}; % TODO: cite
it is now available in standalone header-only form.
% TODO: seems awkward to put the lineage here; also, we should talk up the use
% of Armadillo as its only dependency, which is nice, and maybe emphasize the
% utility of a header-only library

>>>>

Mathematical optimization is the workhorse of virtually all machine learning
algorithms.  For a given objective function $f(\cdot)$
(which may have a special structure or constraints),
almost all machine learning problems can be boiled down
to the following optimization form:
%
\begin{equation}
\operatorname{argmin} f(x)
\end{equation}

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data. %TODO: cite
Even popular machine learning models such as logistic regression % TODO: cite
have training times mostly dominated by an optimization.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

The ubiquity of optimization in machine learning algorithms highlights the need
for robust and flexible implementations of optimization algorithms.
We present {\tt ensmallen}, a C++ optimization toolkit
that contains a wide variety of optimization techniques for many types of
objective functions.  Through the use of C++ template metaprogramming~\cite{TODO},
{\tt ensmallen} is able to generate efficient code that can help with the
demanding computational needs of many machine learning algorithms.

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe, % TODO: cite
PyTorch, % TODO: cite
and TensorFlow % TODO: cite
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn} % TODO: cite
% TODO: can we find another library to put here too?  Shark maybe?
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt scipy.optimize} % TODO: check and cite
is widely used in the Python community, and MATLAB's {\tt TODO()}
% TODO: what's the name of that MATLAB function? or one of them?
% TODO: cite
has been available and used for many decades.  However, these implementations
are often not suitable for modern machine learning tasks---for instance,
computing the full gradient of the objective function may not be feasible
because there are too many data points.

In this paper, we describe the functionality of {\tt ensmallen} and the types of
problems that it can be applied to.  We discuss the mechanisms by which {\tt ensmallen} is
able to provide both computational efficiency and ease-of-use.
We show a few examples that use the library, as well as empirical performance comparisons
with other optimization libraries.

{\tt ensmallen} is open-source software licensed under the 3-claused BSD license~\cite{TODO},
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{TODO} is used for efficient linear algebra operations.

% Originally, {\tt ensmallen} was developed as part of {\tt mlpack}; % TODO: cite
% it is now available in standalone header-only form.
% TODO: seems awkward to put the lineage here;
%% CS: agreed; i don't see a need to explicitly mention mlpack -- it distracts
%% CS: from the central message
% TODO: also, we should talk up the use
% of Armadillo as its only dependency, which is nice, and maybe emphasize the
% utility of a header-only library
